Draft final report\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}

### 2014-10-29 draftfinalreport.Rnw
revisions

\usepackage{times}
\usepackage{hyperref}
\usepackage{tikz}
%\usepackage{bibunits}
\usepackage{natbib}
\usepackage{graphics}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{eurosym}
\usepackage{todonotes}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{array}
\usepackage{rotating}
\usepackage{threeparttable}
\usepackage{blindtext}
\usepackage{dcolumn}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption2}
\usepackage{fancyhdr}
\newcommand{\vecA}{\mathbf{A}}
\newcommand{\vecB}{\mathbf{B}}
\newcommand{\vecC}{\mathbf{C}}
\newcommand{\vecD}{\mathbf{D}}
\newcommand{\vecE}{\mathbf{E}}
\newcommand{\vecF}{\mathbf{F}}
\newcommand{\vecH}{\mathbf{H}}
\newcommand{\vecI}{\mathbf{I}}
\newcommand{\vecJ}{\mathbf{J}}
\newcommand{\vecK}{\mathbf{K}}
\newcommand{\vecM}{\mathbf{M}}
\newcommand{\vecN}{\mathbf{N}}
\newcommand{\vecP}{\mathbf{P}}
\newcommand{\vecQ}{\mathbf{Q}}
\newcommand{\vecS}{\mathbf{S}}
\newcommand{\vecT}{\mathbf{T}}
\newcommand{\vecW}{\mathbf{W}}
\newcommand{\vecZ}{\mathbf{Z}}
\newcommand{\vecb}{\mathbf{b}}
\newcommand{\vecc}{\mathbf{c}}
\newcommand{\vece}{\mathbf{e}}
\newcommand{\vecf}{\mathbf{f}}
\newcommand{\vecg}{\mathbf{g}}
\newcommand{\vech}{\mathbf{h}}
\newcommand{\veci}{\mathbf{i}}
\newcommand{\vecr}{\mathbf{r}}
\newcommand{\vecu}{\mathbf{u}}
\newcommand{\vecv}{\mathbf{v}}
\newcommand{\vecw}{\mathbf{w}}
\newcommand{\vecx}{\mathbf{x}}
\newcommand{\vecy}{\mathbf{y}}
\newcommand{\vecz}{\mathbf{z}}
\newcommand{\Ach}{\mathcal{H}}
\newcommand{\Ell}{\mathcal{L}}
\newcommand{\En}{\mathcal{N}}
\newcommand{\Pe}{\mathcal{P}}
\newcommand{\Ro}{\mathcal{R}_0}
\newcommand{\Rc}{\mathcal{R}_c}
\newcommand{\Rr}{\mathcal{R}}
\newcommand{\Qu}{\mathcal{Q}}
\newcommand{\Te}{\mathcal{T}}
\newcommand{\Ve}{\mathcal{V}}
\newcommand{\Dub}{\mathcal{W}}
\newcommand{\Zed}{\mathcal{Z}}
\newcommand{\Exp}{\mathbb{E}}
\newcommand{\de}{\mbox{d}}
\newcommand{\wrt}{\,\mbox{d}}
\newcommand{\veczero}{\mathbf{0}}
\newcommand{\vecone}{\mathbf{1}}
\newcommand{\vecalpha}{\mbox{\boldmath$\alpha$}}
\newcommand{\vectheta}{\mbox{\boldmath$\theta$}}
\newcommand{\vecphi}{\mbox{\boldmath$\phi$}}
\newcommand{\vecPhi}{\mbox{\boldmath$\Phi$}}
\newcommand{\vecSigma}{\mbox{\boldmath$\Sigma$}}
\newcommand{\vecimath}{\mbox{\boldmath$\imath$}}
\newcommand{\cosech}{\mbox{cosech}\,}
\newcommand{\sign}{\mbox{sign}\,}
\newcommand{\trace}{\mbox{trace}\,}
\newcommand{\Inf}{\mbox{Inf}\,}

@@

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad. In this draft final report we reviewed the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region. We identified additional analyses that may help understand risk of infection in New Zealand, and present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007, provide updated cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on several alternative situations. We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels. Our key findings were:

\begin{itemize}
\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$750,000.
\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, is well above one. This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist range from 0-53\%, leading to an additional 0--250,000 additional vaccinations.
\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if each measles case in New Zealand continues to infect the same number of secondary cases as we estimate from the current 2013--2014 outbreak, additional vaccination is extremely beneficial, but if outbreak dynamics remain the same as those of the average outbreaks since 1997 and future measles cases are few, there is no benefit to additional vaccination campaigns.
\end{itemize}

@@

The analyses we believe could further inform the understanding of risk from measles infection are:
\begin{itemize}
\item Multivariate modelling to account for confounding within the univariate analyses.
\item Inclusion of multivariate model results with local level immunity data to develop a risk map for New Zealand.
\item Analyses of changing risk factors through time during outbreaks.
\item Modelling transmission chains in the population to understand effective reproduction number.
\item Development of risk maps to understand measles importation.
\end{itemize}

\subsection{Additional risk analyses}
\label{sub:risk_analyses}

In this section we provide a brief description of the work that we believe will help inform the Ministry of Health regarding the understanding of risk from measles. These analyses are intended to build on the analyses already included in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report reviewed above.

\begin{figure}
     \centering
     \includegraphics[width=0.8\textwidth]{incidence_1997_2014.pdf}
     \caption{Measles incidence from 1997 to 2014}
     \label{fig:incidence1997}
\end{figure}

In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand. To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to use quantitative methods to evaluate the risk of measles importation.

\subsection{Additional risk analyses methods}

For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand. We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Note this uses all immigration of foreign nationals, coming for whatever purpose. We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.

<<echo=FALSE>>=
library(xtable)
dpdf<-read.csv("mapdata.csv",header=T)
# output top countries
colnames(dpdf)[1]<-c("country")
topimmigration <- dpdf[order(-dpdf$immigration),] 
immigration<-topimmigration[1:10,c(1,4)]
topincidence <- dpdf[order(-dpdf$incidence),] 
incidence<-topincidence[1:10,c(1,2)]
incidence[,2]<-round(incidence[,2],0)
topvaccine <- dpdf[order(dpdf$cover),] 
vaccinecover<-topvaccine[1:10,c(1,3)]
toprisk <- dpdf[order(-dpdf$risk),] 
risk<-(toprisk[1:10,c(1,5)])
risk[,2]<-round(risk[,2],0)
@

\begin{table}
\caption{Immigration numbers to New Zealand (2012)}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(immigration, file="", table.env=FALSE,rowname=NULL)
@
\label{table:immigration12}
\end{table}

%%%%%%%%%%%%%%%%%%%
% INSERT NZ Travelers here
%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%
% INSERT Total Travelers here
%%%%%%%%%%%%%%%%%%

\begin{table}
\caption{Risk of measles importation to New Zealand in 2012}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(risk, file="", table.env=FALSE,rowname=NULL)
@
\label{table:risk12}
\end{table}

%%%%%%%%%%%%%%%%%%%
% INSERT NZ risk here
%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%
% INSERT Total risk here
%%%%%%%%%%%%%%%%%%
\subsection{Risk analyses methods}
We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014. Initial analyses of those ESR data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors. Specifically we required $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report. These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago. We used these denominator data to determine if there were interactions among specific age categories, Prioritized Ethnicities, and NZDep that might exist among cases allowing better understanding of risk of measles infection.

@@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(mfrow=c(1,1))
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="",cex.lab=2)
hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=1.8)
@
\end{center}
\caption{Age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
\label{fig:ageinyears}
\end{figure}

@@

\begin{table}\tiny
\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(ttyr, file="", table.env=FALSE,rowname=NULL)
@
\label{table:case}
\end{table}

@@

\begin{table}\tiny
\caption{Numbers of measles cases, population sizes and per capita rates of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(PerCap, file="", table.env=FALSE,rowname=NULL)
@
\label{table:percap}
\end{table}

@@

@@

<<echo=FALSE>>=
# haven't had time to sort out which packages I need
library(maps)       # Provides functions that let us plot the maps
library(mapdata)    # Contains the hi-resolution points that mark out the countries.
library(rworldmap)
library(adehabitat)
library(rgdal)
library(raster)
library(maps)
library(maptools)
library(sp)
#library(SDMTools)
library(ggplot2)
require(rgeos)
library(ggmap)
library(mapdata)
#gpclibPermit()
library(mapproj)
library(plyr)
library(grid)
library(gridExtra)
#gpclibPermit()
library(utils)
library(rgdal)
#getwd()
#setwd("~/data")
#unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
load("world.ggmap.Rda")
n <- length(unique(world.ggmap$id))
id = unique(world.ggmap$id)
id<-as.data.frame(id)
dpdf<-read.csv("mapdata.csv",header=T)
# output top countries

@@

np4<-ggplot(dpdf, aes(map_id = id)) +
  geom_map(aes(fill = risk), map =world.ggmap) +
  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
   xlab("longitude") +
  ylab("latitude") +
  theme(text = element_text(size=30))+
  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
#np4
@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
print(np1)
@
\end{center}
\caption{Immigration to New Zealand 2012}
\label{fig:immigration12}
\end{figure}

%%%%%%%%%%%%%%%%%%%
% INSERT NZ Travellers here
%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%
% INSERT Total Travelers here
%%%%%%%%%%%%%%%%%%

@@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
print(np4)
@
\end{center}
\caption{Risk map from measles incidence and immigration rates}
\label{fig:risk12}
\end{figure}

%%%%%%%%%%%%%%%%%%%
% INSERT NZ risk here
%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%
% INSERT Total risk here
%%%%%%%%%%%%%%%%%%

%%%%%%%%%%% REVISE FIGURE %%%%%%%%%%%
\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE>>=
change<-read.csv("change.csv",header=T)
plot(change[18:25,1],change[18:25,4],type="l",xlab="Year",ylab="",ylim=c(0,max(change[,4])),
     bty="n")
ylim <- range(c(0,max(change[,4])))
xlim <- range(change[,1])
par(fig = c(.6, 1, 0.6, 1), mar=c(0,0,0,0), new=TRUE)
plot(change[,1],change[,4],type="l",xlab="Year",ylab="",
     ylim=c(0,max(change[,4])),bg="grey",add=T,bty="n")
@
\end{center}
\caption{Trend in global per capita measles incidence}
\label{fig:trendincidence}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE>>=
ttimes<-read.csv("times.csv",header=T)
lmres<-lm(res~time,data=ttimes)
#summary(lmres)
plot(ttimes$res,xaxt="n",type="l",xlab="Date",ylab="")
axis(1, at=1:71, labels=ttimes$mydate) 
abline(lmres,col="red",
       lty=2)
acf(ttimes$res,lag.max = length(ttimes$res),plot=T,main="")
@
\end{center}
\caption{Trend in total immigrations into New Zealand}
\label{fig:trendimmigration}
\end{figure}

%%%%%%%%% ADD NAIVE ## BY DHB HERE

>>>>

\usepackage{times}
\usepackage{hyperref}
\usepackage{tikz}
%\usepackage{bibunits}
\usepackage{natbib}
\usepackage{graphics}
\usepackage{subfig}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{eurosym}
\usepackage{todonotes}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{array}
\usepackage{rotating}
\usepackage{threeparttable}
\usepackage{blindtext}
\usepackage{dcolumn}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption2}
\usepackage{caption}
\usepackage{fancyhdr}
\usepackage{placeins}
\usepackage{framed}
\newcommand{\vecA}{\mathbf{A}}
\newcommand{\vecB}{\mathbf{B}}
\newcommand{\vecC}{\mathbf{C}}
\newcommand{\vecD}{\mathbf{D}}
\newcommand{\vecE}{\mathbf{E}}
\newcommand{\vecF}{\mathbf{F}}
\newcommand{\vecH}{\mathbf{H}}
\newcommand{\vecI}{\mathbf{I}}
\newcommand{\vecJ}{\mathbf{J}}
\newcommand{\vecK}{\mathbf{K}}
\newcommand{\vecM}{\mathbf{M}}
\newcommand{\vecN}{\mathbf{N}}
\newcommand{\vecP}{\mathbf{P}}
\newcommand{\vecQ}{\mathbf{Q}}
\newcommand{\vecS}{\mathbf{S}}
\newcommand{\vecT}{\mathbf{T}}
\newcommand{\vecW}{\mathbf{W}}
\newcommand{\vecZ}{\mathbf{Z}}
\newcommand{\vecb}{\mathbf{b}}
\newcommand{\vecc}{\mathbf{c}}
\newcommand{\vece}{\mathbf{e}}
\newcommand{\vecf}{\mathbf{f}}
\newcommand{\vecg}{\mathbf{g}}
\newcommand{\vech}{\mathbf{h}}
\newcommand{\veci}{\mathbf{i}}
\newcommand{\vecr}{\mathbf{r}}
\newcommand{\vecu}{\mathbf{u}}
\newcommand{\vecv}{\mathbf{v}}
\newcommand{\vecw}{\mathbf{w}}
\newcommand{\vecx}{\mathbf{x}}
\newcommand{\vecy}{\mathbf{y}}
\newcommand{\vecz}{\mathbf{z}}
\newcommand{\Ach}{\mathcal{H}}
\newcommand{\Ell}{\mathcal{L}}
\newcommand{\En}{\mathcal{N}}
\newcommand{\Pe}{\mathcal{P}}
\newcommand{\Ro}{\mathcal{R}_0}
\newcommand{\Rc}{\mathcal{R}_c}
\newcommand{\Rr}{\mathcal{R}}
\newcommand{\Qu}{\mathcal{Q}}
\newcommand{\Te}{\mathcal{T}}
\newcommand{\Ve}{\mathcal{V}}
\newcommand{\Dub}{\mathcal{W}}
\newcommand{\Zed}{\mathcal{Z}}
\newcommand{\Exp}{\mathbb{E}}
\newcommand{\de}{\mbox{d}}
\newcommand{\wrt}{\,\mbox{d}}
\newcommand{\veczero}{\mathbf{0}}
\newcommand{\vecone}{\mathbf{1}}
\newcommand{\vecalpha}{\mbox{\boldmath$\alpha$}}
\newcommand{\vectheta}{\mbox{\boldmath$\theta$}}
\newcommand{\vecphi}{\mbox{\boldmath$\phi$}}
\newcommand{\vecPhi}{\mbox{\boldmath$\Phi$}}
\newcommand{\vecSigma}{\mbox{\boldmath$\Sigma$}}
\newcommand{\vecimath}{\mbox{\boldmath$\imath$}}
\newcommand{\cosech}{\mbox{cosech}\,}
\newcommand{\sign}{\mbox{sign}\,}
\newcommand{\trace}{\mbox{trace}\,}
\newcommand{\Inf}{\mbox{Inf}\,}

@@

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, yet significant outbreaks of measles after measles introductions from abroad. In this draft final report we review the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region. We identified additional analyses that may help understand risk of infection in New Zealand, and present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007. We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on the numbers of naive people at the District Health Board (DHB) and national level. We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels. Our key findings were:

\begin{itemize}
\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
\item Peak overall travel rates, and thus presumably risk from measles importation, is in December. However, New Zealander and immigrant or non--New Zealander travelling is otherwise out of phase, with peak travel for New Zealanders during the winter, and non--New Zealanders summer.
\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, as analysed from data from April, is well above one. This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist is approximately 28\%, leading to an additional 131,500 vaccinations.
\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$809,000.
\item The benefit--cost ratio analyses suggest additional vaccination is extremely beneficial financially.
\end{itemize}

@@

\section{Additional risk analyses}
\label{sub:risk_analyses}

In this section we provide work that we believe will help inform the Ministry of Health regarding the understanding of risk from measles. These analyses are intended to build on the analyses already included in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report reviewed above.

We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand, and descriptive analyses of risk of infection due to previous vaccination history, current rates of immunity within the population. In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand. To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel.

\begin{figure}
     \centering
     \includegraphics[width=1\textwidth]{incidence_1997_2014.pdf}
     \caption{Measles incidence from 1997 to 2014}
     \label{fig:incidence1997}
\end{figure}

%%%%%%%%%%%%%%%%%%
\subsection{Risk of measles infection in New Zealand analyses methods}
We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014. Initial analyses of those ESR data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors. Specifically we required $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report. These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago. We used these denominator data to determine if there were interactions among specific age categories, Prioritized Ethnicities, and NZDep that might exist among cases allowing better understanding of risk of measles infection.

@@

\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width = 15>>=
par(mfrow=c(1,1))
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="",cex.lab=2)
hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=1.8)
@
\end{center}
\caption{Age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
\label{fig:ageinyears}
\end{figure}

@@

\begin{table}[hbtp]
\footnotesize
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(ttyr, file="",table.env=FALSE,rowname=NULL)
@
\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
\label{table:case}
\end{table}

@@

\begin{table}[hbtp]
\footnotesize
\caption{Numbers of measles cases, population sizes and per capita rates of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(PerCap, file="", table.env=FALSE,rowname=NULL)
@
\label{table:percap}
\end{table}

@@

\subsection{Measles importation risk}

For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel data by country and year (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) to measure human movement to and from New Zealand. We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports. We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from. We used simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation. We use the data from 2012 because this year had the most complete WHO measles data and yet was most recent, thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.

%%%%%%%%%%%%%%%%%%
<<echo=FALSE>>=
library(xtable)
totdpdf<-read.csv("mapdata_tot.csv",header=T)
# output top countries
colnames(totdpdf)[1]<-c("country")
toptot <- totdpdf[order(-totdpdf$immigration),] 
immigration<-toptot[1:30,c(1,4)]
topincidence <- totdpdf[order(-totdpdf$incidence),] 
incidence<-topincidence[1:30,c(1,2)]
incidence[,2]<-round(incidence[,2],0)
topvaccine <- totdpdf[order(totdpdf$cover),] 
vaccinecover<-topvaccine[1:30,c(1,3)]
toprisk <- totdpdf[order(-totdpdf$risk),] 
risk<-(toprisk[1:30,c(1,5)])
risk[,2]<-round(risk[,2],0)
@

%%%%%%%%%%%%%%%%%%%%%

\begin{table}
\caption{Total New Zealand traveller numbers by country (New Zealand nationals and immigrants, 2012)}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(immigration, file="", table.env=FALSE,rowname=NULL)
@
\label{table:immigration12}
\end{table}

%%%%%%%%%%%%%%%%%%%
<<echo=FALSE>>=
library(xtable)
nzdpdf<-read.csv("mapdata_nzers.csv",header=T)
# output top countries
colnames(nzdpdf)[1]<-c("country")
topnz <- nzdpdf[order(-nzdpdf$immigration),] 
nz<-topnz[1:30,c(1,4)]
topnzrisk <- nzdpdf[order(-nzdpdf$risk),] 
nzrisk<-(topnzrisk[1:30,c(1,5)])
nzrisk[,2]<-round(nzrisk[,2],0)
@

\begin{table}
\caption{New Zealander travel numbers (2012)}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(nz, file="", table.env=FALSE,rowname=NULL)
@
\label{table:nztravel12}
\end{table}

%%%%%%%%%%%%%%%%%%%%%
<<echo=FALSE>>=
library(xtable)
imdpdf<-read.csv("mapdata_immigration.csv",header=T)
# output top countries
colnames(imdpdf)[1]<-c("country")
topim <- imdpdf[order(-imdpdf$immigration),] 
im<-topim[1:30,c(1,4)]
topimrisk <- imdpdf[order(-imdpdf$risk),] 
imrisk<-(topimrisk[1:30,c(1,5)])
imrisk[,2]<-round(imrisk[,2],0)
@

\begin{table}
\caption{Non-New Zealander travel and immigration numbers (2012)}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(im, file="", table.env=FALSE,rowname=NULL)
@
\label{table:imtravel12}
\end{table}

%%%%%%%%%%%%%%%%%%%%%

\begin{table}
\caption{Risk of measles importation to New Zealand due to New Zealand travellers in 2012}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(nzrisk, file="", table.env=FALSE,rowname=NULL)
@
\label{table:nzrisk12}
\end{table}

\begin{table}
\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(imrisk, file="", table.env=FALSE,rowname=NULL)
@
\label{table:imrisk12}
\end{table}

\begin{table}
\caption{Risk of measles importation to New Zealand in 2012, all travellers}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(risk, file="", table.env=FALSE,rowname=NULL)
@
\label{table:risk12}
\end{table}

%%%%%%%%%%%%%%%%%%

@@

<<echo=FALSE>>=
# haven't had time to sort out which packages I need
library(maps)       # Provides functions that let us plot the maps
library(mapdata)    # Contains the hi-resolution points that mark out the countries.
library(rworldmap)
library(adehabitat)
library(rgdal)
library(raster)
library(maps)
library(maptools)
library(sp)
#library(SDMTools)
library(ggplot2)
require(rgeos)
library(ggmap)
library(mapdata)
#gpclibPermit()
library(mapproj)
library(plyr)
library(grid)
library(gridExtra)
#gpclibPermit()
library(utils)
library(rgdal)
#getwd()
#setwd("~/data")
#unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
load("world.ggmap.Rda")
n <- length(unique(world.ggmap$id))
id = unique(world.ggmap$id)
id<-as.data.frame(id)
dpdf<-read.csv("mapdata_tot.csv",header=T)
# output top countries

@@

np4<-ggplot(dpdf, aes(map_id = id)) +
  geom_map(aes(fill = risk), map =world.ggmap) +
  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
   xlab("longitude") +
  ylab("latitude") +
  theme(text = element_text(size=30))+
  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
#np4
# output top countries
dpdf<-read.csv("mapdata_immigration.csv",header=T)
imnp1<-ggplot(dpdf, aes(map_id = id))  +
  geom_map(aes(fill = immigration), map =world.ggmap) +
  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
   xlab("longitude") +
  ylab("latitude") +
   theme(text = element_text(size=30))+
  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")

imnp4<-ggplot(dpdf, aes(map_id = id)) +
  geom_map(aes(fill = risk), map =world.ggmap) +
  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
   xlab("longitude") +
  ylab("latitude") +
  theme(text = element_text(size=30))+
  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
#np4
dpdf<-read.csv("mapdata_nzers.csv",header=T)
nznp1<-ggplot(dpdf, aes(map_id = id))  +
  geom_map(aes(fill = immigration), map =world.ggmap) +
  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
   xlab("longitude") +
  ylab("latitude") +
   theme(text = element_text(size=30))+
  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")

nznp4<-ggplot(dpdf, aes(map_id = id)) +
  geom_map(aes(fill = risk), map =world.ggmap) +
  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
   xlab("longitude") +
  ylab("latitude") +
  theme(text = element_text(size=30))+
  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
#np4

@

@@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
print(np1)
@
\end{center}
\caption{Total international travel, 2012 }
\label{fig:travel12}
\end{figure}

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
print(nznp1)
@
\end{center}
\caption{New Zealander international travel, 2012 }
\label{fig:nztravel12}
\end{figure}

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
print(imnp1)
@
\end{center}
\caption{Non-New Zealander international travel and immigration, 2012 }
\label{fig:immigration12}
\end{figure}

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
print(np4)
@
\end{center}
\caption{Risk map for measles importation, 2012 }
\label{fig:risk12}
\end{figure}

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
print(nznp4)
@
\end{center}
\caption{Risk map for measles importation from New Zealander international travel, 2012 }
\label{fig:nzrisk12}
\end{figure}

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
print(imnp4)
@
\end{center}
\caption{Risk map for measles importation from non-New Zealander international travel and immigration, 2012 }
\label{fig:imrisk12}
\end{figure}
%%%%%%%%%%% REVISE FIGURE %%%%%%%%%%%
\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE>>=
change<-read.csv("change.csv",header=T)
plot(change[18:25,1],change[18:25,4],type="l",xlab="Year",ylab="",ylim=c(0,max(change[,4])),
     bty="n")
ylim <- range(c(0,max(change[,4])))
xlim <- range(change[,1])
par(fig = c(.6, 1, 0.6, 1), mar=c(0,0,0,0), new=TRUE)
plot(change[,1],change[,4],type="l",xlab="Year",ylab="",
     ylim=c(0,max(change[,4])),bg="grey",add=T,bty="n")
@
\end{center}
\caption{Trend in global per capita measles incidence}
\label{fig:trendincidence}
\end{figure}

\begin{figure}
     \begin{center}
     \includegraphics[width=1\textwidth]{nzers.pdf}
     \end{center}
     \caption{Trends in international travel}
     \label{fig:travel}
\end{figure}

@@

\begin{figure}
     \begin{center}
     \includegraphics[width=1.2\textwidth]{naive_map.pdf}
     \end{center}
     \caption{Numbers of naive individuals per District Health Board}
     \label{fig:naive_map}
\end{figure}
